Tool-driven Revolutions in Archaeological Science

### 2018-09-03 analysis/paper/jcaa-paper.Rmd
working on the text

---
title: "Reproducible research in archaeology using R & rrtools"
author:
  - Sophie C. Schmidt
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
---

<!-- This is the format for text comments that will be ignored during renderings. Do not put R code in these comments because it will not be ignored. -->

@@

# Narrative to accompany the slides for our talk

These are our notes for preparing our slides for our CAA2018 presentation in the session "R as an archaeological tool: current state and directions" (https://github.com/MartinHinz/isaak_conf_sessions/tree/master/caa_2018_tuebingen)

## Slide: Title slide

*Reproducible research in archaeology using R & rrtools*: In this talk, we will show  how archaeologists are changing their data analysis practices to make their research more reproducible and open. We will introduce the rrtools package, which is our specific contribution to help with the challenge of making reproducible research in archaeology easier.

@@

## Slide: Motivation & Outline

 _Motivation_

Our motivation for this topic arises from a profound change that we have been observing in how researchers analyse their data.

Increasingly, we are seeing researchers publishing research that uses a programming language to calculate their models. Similarly, we see journal articles accompanied by scripts of programming language so that readers can study the details of the data analysis methods.

This is an important change away from using mouse-driven, point-and-click software for data analysis. When archaeologists use software such as SPSS and Past (as you see in our screenshots here), it is very difficult for them to share the details of their calculations. When readers cannot see the details of an analysis, it is difficult for them to be confident in the results described in an article. Also, for other researchers it may be near impossible to implement the described methods in their own work as they lack the details on how to come to the desired conclusions. Thus, using scripting language offers the chance of accelerating archaeological research by providing a way to retracing and checking published analyses as well as explaining methods more efficiently. These are the main advantages of creating reproducible research, which we would like to encourage.

_Outline_

In this presentation we will briefly review some data showing change in the use of programming languages in the sciences generally, and in archaeology in particular. We focus on the R programming language, because this is by far the most widely used language in archaeology.

Secondly, we will review how R is used by archaeologists, to show the potential that using a programming language has for opening and accelerating archaeological research.

Finally, we will briefly introduce our R package, rrtools, which is designed to simplify and standardise the process of using R for archaeology, and making research more open and reproducible.

## Slide: Current uses of R [signpost slide only]
So, let's start with the current uses of R.

>>>>

---
title: "Tool-driven Revolutions in Archaeological Science"
author:
  - Sophie C. Schmidt
  - Ben Marwick
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::html_document2:
      number_sections: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
abstract: "There is an argument in philosophy of science that revolutions in science are either ideas-driven or tool-driven. We explore this debate in light of recent efforts by many scientific disciplines to embrace methods to improve the reproducibility of their research. One of the most profound changes driven by this concern for reproducibility and transparency is from analysing data using tools dependent on pointing-and-clicking with a mouse in commercial software, to a new set of tools based on writing scripts in free and open source programming languages. We review bibliometric evidence for this change in the sciences generally and in archaeology in particular to test if the adoption of these new tools may be considered a revolution or some other process. We evaluate how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches."
---

@@

# Introduction

In this paper we investigate recent changes in the way archaeology is being done, especially the computational aspects.

[need some more in here]

# Revolutions in Science: Ideas-driven or Tool-driven?

There is an argument among philosophers of science that revolutions in science are either ideas-driven or tool-driven. One of the most widely-known models of ideas-driven change in science is Kuhn's [-@Kuhn_1962] concept of a paradigm shift. In brief, this model proposes that the history of science consists of long periods of 'normal science' punctuated by short episodes of revolution. Normal science was described by Kuhn as routine day-to-day research focused on 'puzzle solving.' Revolutions in science occur when anomalous observations culminate in changes that replace one paradigm with another.

... Galison

# Tools Emerging from the Reproducibility Crisis

We explore this debate in light of recent efforts by many scientific disciplines to embrace methods to improve the reproducibility of their research. One of the most profound changes driven by this concern for reproducibility and transparency is from analysing data using tools dependent on pointing-and-clicking with a mouse in commercial software, to a new set of tools based on writing scripts in free and open source programming languages...

# Tool-adoption in archaeology and elsewhere

We review bibliometric evidence for this change in the sciences generally and in archaeology in particular to test if the adoption of these new tools may be considered a revolution or some other process. We evaluate how computational approaches to improving reproducibility and transparency in archaeology are mediated and transformed by digital approaches.

# Discussion

rrtools... huskydown

special role of archaeology here, since we are concerned about past behaviours, we have a special interest in research leaving  a visible trace of our work

# Conclusion

@@
